Practical Optional Types for Clojure

### 2015-02-26 formal-model.tex
use formalism in judgements

The first insight about occurrence typing is that
propositions can be used to capture type information about our programs,
in particular by referencing components or subcomponents of the runtime environment
and relating them to types.
\emph{Type Propositions} make assertions like ``variable \clj{x} is of type \clj{Number}'' or
``variable \clj{x} is not \clj{nil}''---
in our logical system we write these as
{\isprop{\clj{Number}}{\x{}}}
and {\notprop{\clj{nil}}{\x{}}}. 
The other propositions are standard: the implication, conjunction,
disjunction, trivial (\topprop{}) and impossible (\botprop{}) propositions
(figure~\ref{main:figure:termsyntax}).

@@

The second insight is that we can replace the traditional 
representation of a
type environment of a map from variables to types
with a sequence of propositions, written \propenv{}. 
Instead of mapping \clj{x} to
the type \clj{Number}, we instead use the proposition {\isprop{\clj{Number}}{\x{}}}.

The third insight is that we
can use a logic system to learn new information about our programs.
The key component of the logic system L-Update uses type propositions
to derive new propositions.
$$
  {\LUpdate}
$$
For example, with L-Update we can use the knowledge of
\inpropenv{\propenv{}}{\isprop{\clj{(U nil Number)}}{\x{}}}
and 
\inpropenv{\propenv{}}{\notprop{\clj{nil}}{\x{}}}
to derive \inpropenv{\propenv{}}{\isprop{\clj{Number}}{\x{}}}.
(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
We cover \updateliteral{} in section~\ref{sec:updateformalsection}.

The fourth insight is a consequence of
programmers in dynamic languages
using implicit type-based reasoning based on the result of
conditional tests.
For instance, example~\ref{example:conditionalflow}
only utilises \clj{x} once
the programmer is convinced it is safe to do so based whether
\clj{(number? x)}
is 
a true or false value. The takeaway is that every expression has
two extra pieces of information relevant to the logic system: a `then' proposition
for when it reduces to a true value, and an `else' proposition
when it reduces to a false value---for \clj{(number? x)}
the then proposition is {\isprop{\clj{Number}}{\x{}}} and 
the else proposition is {\notprop{\clj{Number}}{\x{}}}.
%\ref{main:figure:typingrules}

@@

Types include the top type, \emph{untagged} unions, functions, singleton types
and class instances. 
We write the class literal \clj{Boolean} as \Boolean{} and \clj{Keyword} as \Keyword{}.
The type \Value{\Keyword} is inhabited by \clj{Keyword} and \clj{:a} is of type \Keyword{}.
We abbreviate \EmptyUnion{} as \Bot{}, {\ValueNil} as \Nil{}, 
{\ValueTrue} as \True and {\ValueFalse} as {\False}.
Function types contain \emph{latent}(terminology from~\cite{Lucassen88polymorphiceffect}) propositions and object, which, along with the return type,
may refer to the parameter identifier. They are latent because they are instantiated with the
actual object of the argument in applications before they are used in the proposition environment.

@@

We can now demonstrate how to check example~\ref{example:conditionalflow}.
We abbreviate \clj{Num} as \clj{N},  \clj{Boolean} as \clj{B} and \clj{number?} as \clj{n?}.
The test is checked with T-App, 
$$
\judgement{\isprop{\clj{(U nil N)}}{\x{}}}{\hastype{\clj{(n? x)}}{\clj{B}}}{\filterset{\isprop{\clj{N}}{\x{}}}{\notprop{\clj{N}}{\x{}}}}{\emptyobject{}}
$$
\clj{n?} has type
$$
{\ArrowOne{\x{}}{\clj{Any}}{\clj{B}}
        {\filterset{\isprop{\clj{N}}{\x{}}}{\notprop{\clj{N}}{\x{}}}}{\emptyobject{}}}
$$
By T-Local.
$$
\judgement{\isprop{\clj{N}}{\x{}}}{\hastype{\clj{x}}{\clj{N}}}{\filterset{\notprop{\clj{(U nil false)}}{\x{}}}{\isprop{\clj{(U nil false)}}{\x{}}}}{\x{}}
$$
Now we check both branches.
$$
\judgement{\isprop{\clj{N}}{\x{}}}{\hastype{\clj{(inc x)}}{\clj{N}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$
$$
\judgement{\notprop{\clj{N}}{\x{}}}{\hastype{\clj{0}}{\clj{N}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$
\clj{inc} has type
$$
{\ArrowOne{\x{}}{\clj{N}}{\clj{N}}
        {\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}}
$$
We can now check the conditional with T-If.
$$
\judgement{\isprop{\clj{N}}{\x{}}}{\hastype{\clj{(if (n? x) (inc x) 0)}}{\clj{N}}}{\filterset{\orprop{\isprop{\clj{N}}{\x{}}}{\topprop{}}}{\orprop{\notprop{\clj{N}}{\x{}}}{\topprop{}}}}{\emptyobject{}}
$$
Finally the function can be checked with T-Abs
$$
\judgement{}{\hastype{\clj{(fn ...)}}
                                             {\ArrowOne{\x{}}{\clj{(U nil N)}}{\clj{N}}
        {\filterset{\orprop{\isprop{\clj{N}}{\x{}}}{\topprop{}}}{\orprop{\notprop{\clj{N}}{\x{}}}{\topprop{}}}}{\emptyobject{}}}}
  {\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
$$

The operational semantics for our language are standard, except \nil{} is a false value:

@@

Now we can check example~\ref{example:doexception}.
We check the test with T-App.
$$
\judgement{\isprop{\clj{(U nil N)}}{\x{}}}{\hastype{\clj{(n? x)}}{\clj{B}}}{\filterset{\isprop{\clj{N}}{\x{}}}{\notprop{\clj{N}}{\x{}}}}{\emptyobject{}}
$$
By T-Nil and subsumption we derive the same information in both propositions.
$$
\judgement{\isprop{\clj{N}}{\x{}}}{\hastype{\clj{nil}}{\clj{nil}}}{\filterset{\isprop{\clj{N}}{\x{}}}{\notprop{\clj{N}}{\x{}}}}{\emptyobject{}}
$$
By T-Error we can assume anything about the return of an error.
$$
\judgement{\notprop{\clj{N}}{\x{}}}{\hastype{\errorvalv{}}{\Bot}}{\filterset{\botprop{}}{\botprop{}}}{\emptyobject{}}
$$
Using the above as premises to T-If we learn {\orprop{\isprop{\clj{Num}}{\x{}}}{\botprop{}}}
in both propositions.
$$
\judgement{\isprop{\clj{(U nil N)}}{\x{}}}{\hastype{\clj{(if (n? x) nil}\ {\errorvalv{}{\clj{)}}}}{\clj{B}}}
          {\filterset{\orprop{\isprop{\clj{N}}{\x{}}}{\botprop{}}}{\orprop{\notprop{\clj{N}}{\x{}}}{\botprop{}}}}{\emptyobject{}}
$$
By T-Local we know \clj{x} is a \clj{N}.
$$
\judgement{\orprop{\isprop{\clj{N}}{\x{}}}{\botprop{}}}{\hastype{\clj{x}}{\clj{N}}}{\filterset{\notprop{\clj{(U nil false)}}{\x{}}}{\isprop{\clj{(U nil false)}}{\x{}}}}{\x{}}
$$
By T-App,
\clj{inc} requires a \clj{Num} argument which we just proved.
$$
\judgement{\orprop{\isprop{\clj{N}}{\x{}}}{\botprop{}}}{\hastype{\clj{(inc x)}}{\clj{N}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$
Finally we conclude with T-Do using the conditional and increment expressions as premises.
$$
\judgement{\isprop{\clj{(U nil N)}}{\x{}}}{\hastype{\clj{(do (if (n? x) nil}\ {\errorvalv{}}{\clj{) (inc x))}}}{\clj{N}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$

@@

    {\TAssoc}
  \end{mathpar}
  \caption{HMap Terms, Types and Typing Rules}
  \label{main:figure:hmapsyntax}
\end{figure}

\begin{figure}
  \footnotesize
  \begin{mathpar}
    {\BAssoc}
    {\BGet}
    {\BGetMissing}
  \end{mathpar}
  \begin{mathpar}
    \HMapsubtyping{}
  \end{mathpar}
  \caption{HMap Semantics and Subtyping rules}
  \label{main:figure:hmapsem}
\end{figure}

@@

The expressions \clj{(get m :a)} and \clj{(:a m)} are semantically identical, though
we only model the former to avoid the added complexity of keywords being functions.
To simplify presentation, we only provide syntax for the empty map literal and
restrict lookup and extension to keyword keys. The metavariable \mapval{}
ranges over the runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}}.

Subtyping for HMaps (figure~\ref{main:figure:hmapsem})
designate \MapLiteral{} as a common supertype for all HMaps.
S-HMap says that a HMap is a subtype of another HMap if they agree
on \completenessmeta{}, agree on mandatory entries with subtyping
and at least cover the absent keys of the supertype.
Complete maps are subtypes of partial maps
as long as they agree on the mandatory entries of the partial map via subtyping (S-HMapP).

@@

Figure~\ref{main:figure:hmapsem} contains the semantics for \getliteral{}
and \assocliteral{}.
The B-Get rule evaluates first evaluates the map and the key, asserts
the key must be present in the map and returns the associated key. If
the entry is missing, B-GetMissing says we return \nil{}.

@@

Recall the first insight of occurrence typing---not
only do we want to reason about components of the runtime environment, that
is, variables, but \emph{subcomponents} of the environment.
We have already mentioned paths \path{\pathelem{}}{\x{}}
where \pathelem{} is a sequence of selectors on \x{} called \emph{path elements},
but we have not yet defined what components of a path can be.
\citet{TF10} introduce the path elements \carpe{} and \cdrpe{}
to reason about selector operations on cons cells.
We instead want to reason about HMap lookups.

We introduce our first path element
{\keype{\k{}}}, which represents the operation of looking up
a key \k{}.
We directly relate this to our typing rule T-GetHMap
(figure~\ref{main:figure:hmapsyntax}) by
demonstrating how the first conditional test is checked in example~\ref{example:decleaf}.
First we unfold the destructuring syntax.
\begin{minted}{clojure}
  (fn [m :- Expr]
    (let [op (:op m)]
      (if (= :if (:op m))
        {:op :if, ...}
        (if ...))))
\end{minted}

The \classconst{} function is a constant which returns
the argument's class or \nil{}.

$$
\begin{array}{lrlr}
  \pesyntax{}   &::=& \ldots \alt {\classpe{}}
                &\mbox{Path Elements}
\end{array}
$$

\begin{mathpar}
\constanttypefigure{}
\end{mathpar}

@@

@@

\begin{figure*}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \mininonreflectiveexpsyntax{}
    &\mbox{Expressions}\\

    \tatypesyntax{}\\
    \typehintenvsyntax{}\\
    \classtableallsyntax{}
  \end{altgrammar}
  $$
  \caption{Internal Language Reflection Resolution Extensions}
  \label{main:figure:nonreflectivesyntax}
\end{figure*}

\begin{figure}
  \footnotesize
  \begin{mathpar}
    %\RAbs{}
%
%    \RNewElimRefl{}
%
%    \RMethodElimRefl{}
%
    \RFieldElimRefl{}
%
%    \RLet{}
%

    \RLetHint{}
  \end{mathpar}
\caption{Reflection Elimination %(select rules, figure~\ref{appendix:figure:rewrite} for full rules)
}
\label{main:figure:rewrite}
\end{figure}

@@

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \BField{}

    \BMethod{}
  \end{mathpar}
  \caption{Java Interoperability Semantics}
\end{figure}

\begin{figure*}
  \footnotesize
  \begin{mathpar}
    {\TNewStatic}

    {\TFieldStatic}
    {\TMethodStatic}
  \end{mathpar}
  \caption{Java Interoperability Typing Rules}
  \label{main:figure:javatyping}
\end{figure*}

@@

We now have two languages: an internal language that may contain non-reflective Java calls (figure 
  \ref{main:figure:nonreflectivesyntax}) 
  and an external language that resembles Clojure, where we must rely on type-hints to resolve
  reflective Java calls (figure~\ref{main:figure:reflectivesyntax}).

@@

@@

The definition of \isacompareliteral{} (figure~\ref{main:figure:isaprops}) is deliberately conservative.
The first line considers the case where the object of the left argument
is a non-empty path ending in \classpe{} and the type of the right argument is a singleton class.
The simplest example

@@

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \isapropsfigure{}
  \end{mathpar}

  \center{Takes \emph{object} of the left parameter and \emph{type} of the right parameter.}
  \caption{Proposition Set for \isaliteral{} expressions}
  \label{main:figure:isaprops}
\end{figure}

@@

\begin{figure*}
  \footnotesize
$$
\begin{altgrammar}
  \e{} &::=& \ldots \alt \multimethodexpsyntax{} 
                &\mbox{Expressions} \\
  \v{} &::=& \ldots \alt {\multi {\v{}} {\disptable{}}}
                &\mbox{Values} \\
  \s{}, \t{} &::=& \ldots \alt {\MultiFntype{\t{}}{\t{}}}
                &\mbox{Types} \\

 \disptablesyntax{} \\
\end{altgrammar}
$$
\caption{Multimethod Syntax}
\label{main:figure:mmsyntax}
\end{figure*}

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \TDefMulti{}

    \TDefMethod{}
  \end{mathpar}
  \caption{Multimethod typing rules}
  \label{main:figure:mmtyping}
\end{figure}

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \Multisubtyping{}
  \end{mathpar}
  \caption{Multimethod subtyping rules}
  \label{main:figure:mmsubtype}
\end{figure}

\begin{figure}
  \footnotesize
  \getmethodfigure{}
  \begin{mathpar}
    \BDefMethod{}
    \BDefMulti{}
    \BBetaMulti{}
  \end{mathpar}
  \caption{Multimethod and method resolution semantics}
  \label{main:figure:mmsemantics}
\end{figure}

@@

Multimethod semantics are in figure~\ref{main:figure:mmsemantics}.
B-DefMulti creates a multimethod with a dispatch function and an empty dispatch table.
B-DefMethod returns a new multimethod with an extended dispatch table.
B-BetaMulti invokes the dispatch function with the evaluated argument to get the dispatch value,
which \getmethodliteral{} uses to choose the method to invoke.

The typing rules (figure~\ref{main:figure:mmtyping}) use the multimethod type {\MultiFntype{\s{}}{\t{}}}, 
where \s{} is the interface type for beta reduction, and \t{} is the type for
the dispatch function. T-DefMulti uses the provided interface type and dispatch function
to infer the multimethod type. T-DefMethod uses \isacompareliteral{} to infer the proposition
that must be true if this method is being invoked. The example can be type checked without
changes since we infer \isprop{\Number}{\x{}} and \isprop{\Keyword}{\x{}} for the respective
method bodies. T-App handles beta reduction because multimethods are subtypes of functions
(figure~\ref{main:figure:mmsubtype}).

>>>>

The first insight about occurrence typing is that
propositions can be used to capture type information about our programs,
in particular by referencing components or subcomponents of the runtime environment
and relating them to types.
\emph{Type Propositions} make assertions like ``variable \clj{x} is of type \clj{Number}'' or
``variable \clj{x} is not \clj{nil}''---
in our logical system we write these as
{\isprop{\NumberFull}{\x{}}}
and {\notprop{\Nil{}}{\x{}}}. 
The other propositions are standard: the implication, conjunction,
disjunction, trivial (\topprop{}) and impossible (\botprop{}) propositions
(figure~\ref{main:figure:termsyntax}).

@@

The second insight is that we can replace the traditional 
representation of a
type environment of a map from variables to types
with a sequence of propositions, written \propenv{}. 
Instead of mapping \clj{x} to
the type \clj{Number}, we use the proposition {\isprop{\NumberFull}{\x{}}}.

The third insight is that we
can use a logic system to learn new information about our programs.
The key component of the logic system L-Update uses type propositions
to derive new propositions.
$$
  {\LUpdate}
$$
For example, with L-Update we can use the knowledge of
\inpropenv{\propenv{}}{\isprop{\UnionNilNum}{\x{}}}
and 
\inpropenv{\propenv{}}{\notprop{\nil{}}{\x{}}}
to derive \inpropenv{\propenv{}}{\isprop{\Number}{\x{}}}.
(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
We cover \updateliteral{} in section~\ref{sec:updateformalsection}.

The fourth insight is a consequence of
programmers in dynamic languages
using implicit type-based reasoning based on the result of
conditional tests.
For instance, example~\ref{example:conditionalflow}
only utilises \clj{x} once
the programmer is convinced it is safe to do so based whether
\clj{(number? x)}
is 
a true or false value. The takeaway is that every expression has
two extra pieces of information relevant to the logic system: a `then' proposition
for when it reduces to a true value, and an `else' proposition
when it reduces to a false value---for \clj{(number? x)}
the then proposition is {\isprop{\NumberFull}{\x{}}} and 
the else proposition is {\notprop{\NumberFull}{\x{}}}.
%\ref{main:figure:typingrules}

@@

Types include the top type, \emph{untagged} unions, functions, singleton types
and class instances. 
We write the class literal \Booleanlong{} as \Boolean{} and \Keywordlong{} as \Keyword{}.
The type \Value{\Keyword} is inhabited by the class literal \Keyword{} and \clj{:a} is of type \Keyword{}.
We abbreviate \EmptyUnion{} as \Bot{}, {\ValueNil} as \Nil{}, 
{\ValueTrue} as \True and {\ValueFalse} as {\False}.
Function types contain \emph{latent}(terminology from~\cite{Lucassen88polymorphiceffect}) propositions and object, which, along with the return type,
may refer to the parameter identifier. They are latent because they are instantiated with the
actual object of the argument in applications before they are used in the proposition environment.

@@

We can now demonstrate how to check example~\ref{example:conditionalflow}.
We abbreviate \NumberFull{} as \Number{} and \Booleanlong{} as \Boolean{}.
The test is checked with T-App, 
$$
\judgement{\isprop{\UnionNilNum{}}{\x{}}}{\hastype{\appexp{\numberhuh{}}{\x{}}}{\Boolean}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
$$
\numberhuh{} has type
$$
{\ArrowOne{\x{}}{\Top}{\Boolean}
        {\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}}
$$
By T-Local.
$$
\judgement{\isprop{\Number}{\x{}}}{\hastype{\x{}}{\Number{}}}{\filterset{\notprop{\falsy{}}{\x{}}}{\isprop{\falsy{}}{\x{}}}}{\x{}}
$$
Now we check both branches.
$$
\judgement{\isprop{\Number}{\x{}}}{\hastype{\appexp{\inc{}}{\x{}}}{\Number{}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$
$$
\judgement{\notprop{\Number}{\x{}}}{\hastype{\zeroliteral{}}{\Number}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$
\inc{} has type
$$
{\ArrowOne{\x{}}{\Number}{\Number}
        {\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}}
$$
We can now check the conditional with T-If.
$$
\judgement{\isprop{\Number}{\x{}}}{\hastype{\ifexp{\appexp{\numberhuh{}}{\x{}}}{\appexp{\inc{}}{\x{}}}{\zeroliteral{}}}{\Number}}{\filterset{\orprop{\isprop{\Number}{\x{}}}{\topprop{}}}{\orprop{\notprop{\Number}{\x{}}}{\topprop{}}}}{\emptyobject{}}
$$
Finally the function can be checked with T-Abs
$$
\judgement{}{\hastype{\abs{\x{}}{\UnionNilNum}{\ ...}}
                                             {\ArrowOne{\x{}}{\UnionNilNum}{\Number}
        {\filterset{\orprop{\isprop{\Number}{\x{}}}{\topprop{}}}{\orprop{\notprop{\Number}{\x{}}}{\topprop{}}}}{\emptyobject{}}}}
  {\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
$$

The operational semantics for our language standard defined in big-step style, 
The notable different is \nil{} is a false value, which require changes 
to the semantics of \ifliteral{}:

@@

Now we can check example~\ref{example:doexception}.
We check the test with T-App:
$$
\judgement{\isprop{\Union{\nil{}}{\Number}}{\x{}}}{\hastype{\appexp{\numberhuh}{\x{}}}{\Boolean}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
$$
By T-Nil and subsumption we derive the same information in both propositions.
$$
\judgement{\isprop{\Number}{\x{}}}{\hastype{\nil{}}{\nil{}}}{\filterset{\isprop{\Number}{\x{}}}{\notprop{\Number}{\x{}}}}{\emptyobject{}}
$$
By T-Error we can assume anything about the return of an error.
$$
\judgement{\notprop{\Number}{\x{}}}{\hastype{\errorvalv{}}{\Bot}}{\filterset{\botprop{}}{\botprop{}}}{\emptyobject{}}
$$
Using the above as premises to T-If we learn {\orprop{\isprop{\Number}{\x{}}}{\botprop{}}}
in both propositions.
$$
\judgement{\isprop{\Union{\nil{}}{\Number}}{\x{}}}
          {\hastype{\ifexp{\appexp{\numberhuh{}}{\x{}}}{\nil{}}{\errorvalv{}}}{\Boolean}}
          {\filterset{\orprop{\isprop{\Number}{\x{}}}{\botprop{}}}{\orprop{\notprop{\Number}{\x{}}}{\botprop{}}}}{\emptyobject{}}
$$
By T-Local we know \x{} is a \Number{}.
$$
\judgement{\orprop{\isprop{\Number}{\x{}}}{\botprop{}}}{\hastype{\x{}}{\Number}}{\filterset{\notprop{\falsy{}}{\x{}}}{\isprop{\falsy{}}{\x{}}}}{\x{}}
$$
By T-App,
\inc{} requires a \Number{} argument which we just proved.
$$
\judgement{\orprop{\isprop{\Number}{\x{}}}{\botprop{}}}{\hastype{\appexp{\inc{}}{\x{}}}{\Number}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$
Finally we conclude with T-Do using the conditional and increment expressions as premises.
$$
\judgement{\isprop{\Union{\nil{}}{\Number}}{\x{}}}
          {\hastype{\doexp
                     {\ifexp{\appexp{\numberhuh{}}{\x{}}}
                        {\nil{}}
                        {\errorvalv{}}}
                     {\appexp{\inc{}}{\x{}}}}{\Number}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$

@@

    {\TAssoc}
  \end{mathpar}
  \begin{mathpar}
    {\BAssoc}
    {\BGet}
    {\BGetMissing}
  \end{mathpar}
  \begin{mathpar}
    \HMapsubtyping{}
  \end{mathpar}
  \caption{HMap Terms, Types, Semantics, Subtyping and Typing Rules}
  \label{main:figure:hmapsyntax}
\end{figure}

@@

The expressions \clj{(get m :a)} and \clj{(:a m)} are semantically identical, though
we only model the former to avoid the added complexity of keywords being functions.
To simplify presentation, we only provide syntax for the empty map literal and
restrict lookup and extension to keyword keys. The metavariable \mapval{}
ranges over the runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}},
usually written {\curlymapvaloverrightnoarrow{\k{}}{\v{}}}.

Subtyping for HMaps (figure~\ref{main:figure:hmapsyntax})
designate \MapLiteral{} as a common supertype for all HMaps.
S-HMap says that a HMap is a subtype of another HMap if they agree
on \completenessmeta{}, agree on mandatory entries with subtyping
and at least cover the absent keys of the supertype.
Complete maps are subtypes of partial maps
as long as they agree on the mandatory entries of the partial map via subtyping (S-HMapP).

@@

Figure~\ref{main:figure:hmapsyntax} contains the semantics for \getliteral{}
and \assocliteral{}.
The B-Get rule evaluates first evaluates the map and the key, asserts
the key must be present in the map and returns the associated key. If
the entry is missing, B-GetMissing says we return \nil{}.

@@

Recall the first insight of occurrence typing---not
only do we want to reason about components of the runtime environment, that
is, variables, but \emph{subcomponents} of the environment.
We briefly mentioned paths, written \path{\pathelem{}}{\x{}}, but we concentrated on
reasoning only about the variable \x{}---the notation \pathelem{} is a nested sequence of \emph{path elements}
applied right-to-left to a variable \x{}.
\citet{TF10} introduce the path elements \carpe{} and \cdrpe{}
to reason about selector operations on cons cells.
We instead want to reason about HMap lookups.

We introduce our first path element
{\keype{\k{}}}, which represents the operation of looking up
a key \k{}.
We directly relate this to our typing rule T-GetHMap
(figure~\ref{main:figure:hmapsyntax}) by
checking the then branch of the first conditional test is checked in 
an equivalent version of example~\ref{example:decleaf}.
\begin{minted}{clojure}
  (fn [m :- Expr]
    (if (= (get m :op) :if)
      {:op :if, ...}
      (if ...)))
\end{minted}

We do not specifically support \equivliteral{} in our calculus, 
but on keyword arguments it works identically to \clj{isa?} which we model
in section~\ref{sec:isaformal}.
Intuitively, if {\judgement{\propenv{}}{\hastype{\e{}}{\t{}}}{\filterset{\thenprop{\prop{}}}{\elseprop{\prop{}}}}{\object{}}}
then \equivapp{\e{}}{\makekw{if}} has the proposition set 
$$
{\replacefor{\filterset{\isprop{\Value{\makekw{if}}}{\x{}}}{\notprop{\Value{\makekw{if}}}{\x{}}}}{\object{}}{\x{}}}
$$
where substitution reduces to \topprop{} if \object{} = \emptyobject{}.

By T-Abs, checking the body of the function we have the proposition environment \propenv{} = {\isprop{\Expr{}}{m}}.
By T-Local we have
$$
\judgement{\propenv{}}{\hastype{m}{\Expr{}}}{\filterset{\notprop{\falsy{}}{\x{}}}{\isprop{\falsy{}}{\x{}}}}{m}
$$
and since {\Expr{}} is a union of HMaps with the entry \makekw{op}, we use T-GetHMap to derive
$$
\judgement{\propenv{}}{\hastype{\getexp{m}{\makekw{op}}}{\Keyword}}{\filterset{\topprop{}}{\topprop{}}}{\object{1}}
$$
where \object{1} = {\path{\keype{\makekw{op}}}{m}}. Using our intuitive definition of \equivliteral{} above, we know
$$
\judgement{\propenv{}}{\hastype{\equivapp{\getexp{m}{\makekw{op}}}{\makekw{if}}}{\Boolean}}{\filterset{\isprop{\Value{\makekw{if}}}{\object{1}}}{\notprop{\Value{\makekw{if}}}{\object{1}}}}{\emptyobject{}}
$$
Now we check the then branch with our new assumption.
$$
\judgement{\propenvp{}}{\hastype{\curlymap{{\mapvalentrykwkeynoarrow{op}{\makekw{if}}}, ...}}{\Expr{}}}{...}{...}
$$
We are interested in what L-Update can derive in our new environment,
\propenvp{} = {\isprop{\Expr{}}{m}},{\isprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}.

With the premises
\inpropenv{\propenvp{}}{\isprop{\Expr{}}{m}}
and
\inpropenv{\propenvp{}}{\isprop{\Value{\makekw{if}}}{\path{\keype{\makekw{op}}}{m}}}
we derive
$$
\inpropenv{\propenvp{}}{\isprop{\update{\Expr{}}{\Value{\makekw{if}}}{\keype{\makekw{op}}}}{m}}
$$
which simplifies to 
$$
\inpropenv{\propenvp{}}{\isprop{\HMapp{\mandatoryset{{\mandatoryentrykwnoarrow{op}{\makekw{if}}}, {\mandatoryentrykwnoarrow{test}{\Expr{}}},
                                       {\mandatoryentrykwnoarrow{then}{\Expr{}}},   {\mandatoryentrykwnoarrow{else}{\Expr{}}}}}
                                   {\emptyabsent{}}}{m}}
$$

Our second path element \classpe{} is used in the latent
object of the constant \classconst{} function. Like Clojure's \clj{class}
function \classconst{} returns the argument's class or \nil{}
if passed \nil{}.
$$
\begin{array}{lrlr}
  \pesyntax{}   &::=& \ldots \alt {\classpe{}}
                &\mbox{Path Elements}
\end{array}
$$
\begin{mathpar}
\constanttypefigure{}
\end{mathpar}
The full semantics are given in figure~\ref{main:figure:primitivesem}.
The definition of \updateliteral{} supports various idioms relating to \classpe{}
which we introduce in section~\label{sec:isaformal}.

@@

\begin{figure}[h]
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \mininonreflectiveexpsyntax{}
    \\

    \tatypesyntax{}\\
    \typehintenvsyntax{}\\
    \classtableallsyntax{}
  \end{altgrammar}
  $$
  \begin{mathpar}
    {\TNewStatic}
    {\TFieldStatic}

    {\TMethodStatic}
  \end{mathpar}
  \begin{mathpar}
    \BField{}

    \BMethod{}
  \end{mathpar}
  \begin{mathpar}
    %\RAbs{}
%
%    \RNewElimRefl{}
%
%    \RMethodElimRefl{}
%
    \RFieldElimRefl{}
%
%    \RLet{}
%

    \RLetHint{}
  \end{mathpar}
  \caption{Java Interoperability Syntax, Semantics, Typing Rules and Rewrite Relation}
  \label{main:figure:javatyping}
\end{figure}

In section~\ref{sec:overviewjavainterop} we discussed how Clojure programmers use
type hints to help the compiler eliminate reflective calls.
In this section we introduce our model of Java, including syntax
for unresolved Java

@@

@@

@@

We now have two languages: an internal language that may contain non-reflective Java calls (figure~\ref{main:figure:javatyping}) 
  and an external language that resembles Clojure, where we must rely on type-hints to resolve
  reflective Java calls (figure~\ref{main:figure:reflectivesyntax}).

@@

\label{sec:isaformal}

@@

The definition of \isacompareliteral{} (figure~\ref{main:figure:mmsyntax}) is deliberately conservative.
The first line considers the case where the object of the left argument
is a non-empty path ending in \classpe{} and the type of the right argument is a singleton class.
The simplest example

@@

@@

\begin{figure}
  \footnotesize
$$
\begin{altgrammar}
  \e{} &::=& \ldots \alt {\createmultiexp {\t{}} {\e{}}} &\mbox{Expressions} \\
             &\alt& {\extendmultiexp {\e{}} {\e{}} {\e{}}}
             \alt {\isaapp {\e{}} {\e{}}}\\
  \v{} &::=& \ldots \alt {\multi {\v{}} {\disptable{}}}
                &\mbox{Values} \\
  \s{}, \t{} &::=& \ldots \alt {\MultiFntype{\t{}}{\t{}}}
                &\mbox{Types} \\

 \disptablesyntax{} \\
\end{altgrammar}
$$
  \begin{mathpar}
    \Multisubtyping{}
  \end{mathpar}
  \begin{mathpar}
    \isapropsfigure{}
  \end{mathpar}
  \begin{mathpar}
    \TDefMulti{}

    \TDefMethod{}
  \end{mathpar}
  \getmethodfigure{}
  \begin{mathpar}
    \BDefMethod{}
    \BDefMulti{}
    \BBetaMulti{}
  \end{mathpar}
\caption{Multimethod Syntax, Subtyping, Typing Rules and Semantics}
\label{main:figure:mmsyntax}
\end{figure}

@@

Multimethod semantics are in figure~\ref{main:figure:mmsyntax}.
B-DefMulti creates a multimethod with a dispatch function and an empty dispatch table.
B-DefMethod returns a new multimethod with an extended dispatch table.
B-BetaMulti invokes the dispatch function with the evaluated argument to get the dispatch value,
which \getmethodliteral{} uses to choose the method to invoke.

The typing rules (figure~\ref{main:figure:mmsyntax}) use the multimethod type {\MultiFntype{\s{}}{\t{}}}, 
where \s{} is the interface type for beta reduction, and \t{} is the type for
the dispatch function. T-DefMulti uses the provided interface type and dispatch function
to infer the multimethod type. T-DefMethod uses \isacompareliteral{} to infer the proposition
that must be true if this method is being invoked. The example can be type checked without
changes since we infer \isprop{\Number}{\x{}} and \isprop{\Keyword}{\x{}} for the respective
method bodies. T-App handles beta reduction because multimethods are subtypes of functions
(figure~\ref{main:figure:mmsyntax}).
